\documentclass[11pt]{article}

\usepackage[francais]{babel}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{listings}
\usepackage{epsfig}
\usepackage{hyperref}
\usepackage{gb4e}
\usepackage{enumitem}

\lstset{
  frame=single,
  numbers=left,                    % where to put the line-numbers; possible values are (none, left, right)
  numbersep=5pt,                   % how far the line-numbers are from the code
  literate=  {→}{{$\rightarrow$}}1 {↓}{{$\downarrow$}}1 {↑}{{$\uparrow$}}1 {⇓}{{$\Downarrow$}}1 {⇑}{{$\Uparrow$}}1 {⊂}{{$\subset$}}1 {∪}{{$\cup$}}1 {¬}{{$\lnot$}}1
  {á}{{\'a}}1 {é}{{\'e}}1 {í}{{\'i}}1 {ó}{{\'o}}1 {ú}{{\'u}}1
  {Á}{{\'A}}1 {É}{{\'E}}1 {Í}{{\'I}}1 {Ó}{{\'O}}1 {Ú}{{\'U}}1
  {à}{{\`a}}1 {è}{{\`e}}1 {ì}{{\`i}}1 {ò}{{\`o}}1 {ù}{{\`u}}1
  {À}{{\`A}}1 {È}{{\'E}}1 {Ì}{{\`I}}1 {Ò}{{\`O}}1 {Ù}{{\`U}}1
  {ä}{{\"a}}1 {ë}{{\"e}}1 {ï}{{\"i}}1 {ö}{{\"o}}1 {ü}{{\"u}}1
  {Ä}{{\"A}}1 {Ë}{{\"E}}1 {Ï}{{\"I}}1 {Ö}{{\"O}}1 {Ü}{{\"U}}1
  {â}{{\^a}}1 {ê}{{\^e}}1 {î}{{\^i}}1 {ô}{{\^o}}1 {û}{{\^u}}1
  {Â}{{\^A}}1 {Ê}{{\^E}}1 {Î}{{\^I}}1 {Ô}{{\^O}}1 {Û}{{\^U}}1
  {Ã}{{\~A}}1 {ã}{{\~a}}1 {Õ}{{\~O}}1 {õ}{{\~o}}1
  {œ}{{\oe}}1 {Œ}{{\OE}}1 {æ}{{\ae}}1 {Æ}{{\AE}}1 {ß}{{\ss}}1
  {ű}{{\H{u}}}1 {Ű}{{\H{U}}}1 {ő}{{\H{o}}}1 {Ő}{{\H{O}}}1
  {ç}{{\c c}}1 {Ç}{{\c C}}1 {ø}{{\o}}1 {å}{{\r a}}1 {Å}{{\r A}}1
  {€}{{\euro}}1 {£}{{\pounds}}1 {«}{{\guillemotleft}}1
  {»}{{\guillemotright}}1 {ñ}{{\~n}}1 {Ñ}{{\~N}}1 {¿}{{?`}}1
}


\title{Elvex -- Documentation}
\author{Lionel Clément}
\date{Version 2.13.0 --  (to be updated)}
%November 2019

\begin{document}
\maketitle
\section{Grammar by sample}

An example will familiarize the reader with \texttt{Elvex} before we
go any further. Let us take the case of the text generation of a
nominal phrase in French.

\subsubsection*{Example}

\begin{lstlisting}
NP → det N {
  ↓2 = ↑;
  ↓1 = ⇓2;
  ⇑ = ⇓2;
}

N → adj N {
  [mod:<$Head::$Tail>, $Rest];
  [number:$inhNb] ⊂ ↑;
  ↓1 = [number:$inhNb, gender:$synthGd, $Head];
  ↓2 = [mod:$Tail, $Rest];
  [gender:$synthGd] ⊂ ⇓2;
  ⇑ = ⇓2 ∪ [qual:yes];
}

N → n {
  [mod:NIL];
  ↓1 = ↑;
  ⇑ = ⇓1;
}
\end{lstlisting}

\subsection{Syntagmatic description (\textit{Context-Free Grammar rules})}
 
The first rule of grammar (line 1) defines \texttt{NP} as
consisting of \texttt{det} and \texttt{N} from left to right.

\texttt{det} is a word; it is defined in the lexicon that we will see
later.

Lines 7) and 16) follow the same logic and we have the
context-free grammar:

\begin{lstlisting}[numbers=none]
NP → det N
N → adj N
N → n
\end{lstlisting}

The only sequence we can generate with this simple example is
a noun, preceded by a determiner and possibly one or several adjectives.

Syntagmatic rules are therefore made to describe the order of
components of phrases, sentences and speeches.

\subsection{Operational semantics}

Each rule contains a set of operations. These operations describe the
functional properties of the language, which may or may not depend on
the order of the words.

\subsubsection*{The basics}

We will explain this semantics through the example given to the
beginning of this section before giving details.

\begin{itemize}
\item The lines 2) and 3) express the nature of the links that
  exist between \texttt{NP}, the name \texttt{N} and its determiner
  \texttt{det}; the main one being that agreement in gender and number
  occurs between nouns and their specifier.

  \begin{itemize}
  \item Line 2) assigns the given nominal phrase properties to the
    noun. The attribute of the name is noted $\downarrow2$
    (this is the second term of the right part of the rule), and
    nominal phrase's is noted $\uparrow$.

  \item Line 3) retrieves information from analyses of \texttt{N} to
    assign them to the determiner.  The attribute of the determiner
   is noted $\downarrow1$ (this is the first term of the right part
    of the rule), the attribute from the name \texttt{N}
    is noted $\Downarrow2$.

  \item Line 4) assigns the analyse of \texttt{N} to the synthesized
    attribute \texttt{NP}.

  \end{itemize}
  
  With the two first lines, the noun will be chosen in the lexicon
  according to the concept to be expressed (attribute of the nominal
  phrase $\uparrow$).  A female singular determiner will be chosen if
  the gender of the noun is feminine and the nominal phrase is
  singular.
  
  This mechanism, which makes it possible to articulate lexical
  choices according to the concepts and syntactic properties according
  to lexical properties is at the heart of \texttt{Elvex} and
  implements two types of attributes: inherited attributes (noted
  $\uparrow$, $\downarrow_i$) and synthesized attributes (noted
  $\Uparrow$, $\Downarrow_i$).
  
  Inherited attributes are calculated from left to right of the
  syntagmatic rules. Synthesized attributes are calculated from right
  to left. To put it simply, we can say that the inherited attributes
  convey information from the concept to the lexicon, and that the
  synthesized attributes convey information from the lexicon to higher
  layers of the syntactic analysis. Thus inherited attributes allow you to
  propagate the input to the lexicon to make a lexical choice, whereas
  conversely, synthesized attributes allow to retrieve the lexical information of
  the chosen name according to the concept expressed and to propagate
  this information (grammatical features, selection restrictions,
  phraseological units constraints, etc.)  to the entire nominal
  phrase.
  
\item The lines 8) to 13) allow you to set the gender and
  number agreement of the noun and adjective and also to propagate the
  concepts that produce adjectives. We notice that in French, the
  number is a semantism expressed directly as a sign associated with
  the noun, whereas gender is a grammatical feature given by the
  lexicon. The first will be propagated from the concept to the noun,
  the second will be propagated from the lexicon to the layers
  regulating the agreement.
  

\begin{itemize}
  \item Line 8) is a \textit{guard}, \textit{i. e.} a
    condition to realize
    $\mathtt{N} \rightarrow \mathtt{adj}~ \mathtt{N}$. It expresses that the
    realization of an adjective is conditioned by
    the existence of a list
    \verb#<$Head::$Tail># as the value of an attribute \texttt{mod}
    in the inherited attribute $uparrow$.
    
    \verb#$Head# and
    \verb#$Tail# are two variables that respectively define the
    head and the tail of a list, \textit{i. e.} respectively the
    first item in the list and the rest of the list.
    
    if the condition is verified, 
    \verb#$Head#, \verb#$Tail# and
    \verb#$Rest# will be affected by values given by the
    subsumption\footnote{A feature structure \texttt{A} subsumes
      a feature structure \texttt{B} iif all the features of
      \texttt{B} are present in \texttt{A} with the same values
      or with values that subsume those of the
      values of \texttt{A}. } values of \\\verb#[mod:<$Head::$Tail>, $Rest]#
    by $\uparrow$.
    
  \item Line 9) assign to the variable
    \verb#$inhNb# (\textit{Inherited noun}) the value of the
    \texttt{number} feature of the inherited attribute
    $uparrow$. This allows you to retrieve the number given to the
    nominal phrase.
    
  \item Line 10) assigns to the inherited attribute of the adjective
    the head of the list \verb#$Head# unified\footnote{The unification
      of a feature structure \texttt{A} with a feature structure
      \texttt{B}, is the smallest feature structure \texttt{C} such
      that \texttt{A} subsumes \texttt{C} and \texttt{B} subsumes
      \texttt{C}.} with a structure that gives the synthesized gender
    \verb#$synthGd# (\textit{synthesized gender}) and the inherited
    number \verb#$inhNb#.
    
  \item Line 11) assigns to the inherited attribute of the noun all
    values passed by inheritance (\verb#$Rest#) and adds the feature
    \verb#mod# with the value
    \verb#$Tail#, which is the tail of the list that brings together
    all other modifiers of the noun (other adjective or a proposal for
    example).
    
  \item Line 12) allows you to assign the variable
    \verb#$synthGd# which is the gender synthesized from the noun.
    
  \item Finally line 13) assigns to the synthesized attribute of the
    phrase the synthesized attribute of the noun and adds a feature
    \texttt{qual} which marks that this noun phrase is qualified
    (which makes it possible to restrict others modifications).
    
\end{itemize}

\item The rules on lines 17) to 19) are not specific, except for the
  custody of rule 17) which implements the value \verb#NIL#.

  \begin{itemize}    
  \item Line 17) is a guard which allows us to prevent the inherited
    attribute $\uparrow$ to contain any feature
    \texttt{mod}. \verb#mod:NIL# means that the attribute \verb#mod#
    does not exist in $\uparrow$.
    
  \item Lines 18) and 19) do not require any futher comment.

  \end{itemize}
\end{itemize}

It becomes clear by examining this simple example that the order of
operational rules is not the one given by grammar, but the one
dictated by the availability of operands. Here, a possible order of
application of the rules is 8), 9), 11), 12), 10), and 13). In
the version 2.10 of \texttt{Elvex} this order is calculated on the
fly. In the subsequent versions a dependency graph will be built with
guidelines for developing grammars on this particular aspect.

The advice I can give to develop a grammar is
to write the rules as the synthesized attributes are available.

We see that the syntagmatic rules are not produced
linearly from left to right and from top to bottom, but in such a way that
which is not determined by producing the first available elements.

Before going any further, let us give definitions on the ratings that
  allow us to understand the syntax of the rules.
  
\subsubsection*{Empty rules, optional terms}
 
It is possible to write empty rules, \textit{i.e.}  components that have no
realization in text production. It is also possible to put optional
elements written into brackets signs \verb#[]#.

Example: The following rule gives all possible combinations for clitic
pronouns placed before the verb.

\begin{lstlisting}[numbers=none]
  CLITICS → [clr] [cld] [cla] [cld] [clg] [cll] 
\end{lstlisting}

In French, the dative clitic (\texttt{cld}) is sometimes placed
before, sometimes placed after the accusative clitic (\texttt{cla})
according to the person of the pronoun.

\begin{exe}
\ex \begin{xlist} 
\ex 
\begin{tabbing}
\textit{Jean} xxx \= \textbf{to\_me}(dative) \= \textbf{it}(accusative) \= \textit{donne}\kill
\textit{Jean} \>\textit{\textbf{me}} \>\textit{\textbf{le}} \>\textit{donne}\\
Jean \> \textbf{to\_me}(dative) \> \textbf{it}(accusative) \> gives\\
Jean gives it to me\\
\end{tabbing}

\ex 
\begin{tabbing}
\textit{Jean} xxx \=\textbf{it}(accusative) \=\textbf{to\_him}(dative) \= \textit{donne}\kill
\textit{Jean} \>\textit{\textbf{le}} \>\textit{\textbf{lui}} \>\textit{donne}\\
Jean \>\textbf{it}(accusative) \>\textbf{to\_him}(dative) \>gives\\
Jean gives it to him.
\end{tabbing}
\end{xlist}
\end{exe}
 
According to this rule, the phrase \texttt{CLITICS} can also be empty,
due to the fact that each term is optional.

\subsubsection*{Paradigms}

Each element of a syntagmatic rule can be a list of terms
separated by the character \verb#|# instead of being a single term. In this
case, the product phrase is constructed with one of the terms in the list.

Example:
The following rule allows you to build a sentence without a subject, with a
nominative clitic subject \texttt{cln} (\textit{je}, \textit{tu},
\textit{il}, $\dots$), or with a subject as a nominal group.

\begin{lstlisting}[numbers=none]
S → [cln | NP] VP
\end{lstlisting}

A grammar written by using massively rules with alternatives and
optional terms is equivalent to a very large grammar.  For example,
the rule describing the order of clitic pronouns is equivalent to a
set of 64 rules:
 
\begin{lstlisting}[numbers=none]
  CLITICS → 
  CLITICS → clr 
  CLITICS → cld 
  ...
  CLITICS → clr cld 
  CLITICS → clr cld cla
  ... ...
\end{lstlisting}

However, \texttt{Elvex} never builds these set of rules and remains
effective with very large grammars, even written in doing so.  The use
of alternatives and optional terms is therefore recommended to
describe the order of words, without regard to the combinatorial
problem.

\section{Elvex notations}
 
A syntagmatic rule is written 

$A~ \rightarrow B_1 B_2 \dots B_k <operational~semantics>$
 
Where $B_i$ is written:

\begin{itemize}
\item $[ B_i ]$ if the symbol $B_i$ is optional.
\item $C_1 | C_2 | \dots | C_n$ if $B_i$ is a choice between the $C_i$
\end{itemize}
 
$B_i$, $C_i$ can be lexical categories or symbols of the
grammar.

\subsection{Operational Semantics}

\paragraph{Notations}

Let the rule
$A~ \rightarrow B_1 B_2 \dots B_i \dots B_k$

\begin{itemize}
\item $\uparrow$: refers to the attribute inherited from $A$.
\item $\downarrow_i$: refers to the attribute inherited from $B_i$.
\item $\Uparrow$: refers to the synthesized attribute of $A$.
\item $\Downarrow_i$: refers to the synthesized attribute of $B_i$. 
\end{itemize}

The named variables are noted by an identifier preceded by \og
\$\fg{}, the anonymous variables are noted \og \verb#_#\fg{}, the
constant values are noted by identifiers.

The feature structures are noted $[ f_1, f_2, \dots, f_k ]$
where $ f_i$ is a feature as
\begin{itemize}
\item a linked variable feature
\item an anonymous variable feature
\item \verb#HEAD:<identifier of a lexeme># a feature for a predicate
\item \verb#FORM:"<form>"# a feature for a literal form
\item a pair $attr_i:val_i$
  \begin{itemize}
  \item with $attr_i$ an named attribute
  \item and with $val_i$ the value of an attribute that can be: 
    
    \begin{itemize}
    \item A named variable
    \item An anonymous variable
    \item A nil value (\texttt{NIL})
    \item An atomic constant
    \item A set of atomic constants separated by \verb#|#
    \item A literal constant (number)
    \item A feature structure
    \item A list of feature structures
    \end{itemize}
  \end{itemize}
  
\end{itemize}

The lists are noted
\begin{itemize}
\item $<f_1, f_2, \dots, f_k>$ with $f_i$ as feature structures or lists
\item $<head::tail>$ with $head$ as a feature structure or a list and $tail$, a
  list. $head$ refers to the first item on the list, $tail$ the rest
  of the list.
\end{itemize}

If one wishes to bring an operational semantics to the syntagmatic
rule, one defines $<operational~semantics>$ which is a list of rules
in \{\} brackets.

Each rule is written:

\begin{itemize}

\item $\{<$list of rules $>\}$ A list of rules.

\item \verb#attest <expr>;# Explicit guard where \verb#<expr># is an
  boolean expression that must be checked.

\item \verb#[f_1, feat_2, \dots, f_k];# Guard where the feature
  structure must subsume $\uparrow$. The guard may contain the feature
  \verb#HEAD# which represents the concept to be generated. Thus, a
  concept is not necessarily linked to a lexical entry with
  \texttt{Elvex}, unlike usual text generation systems; it can produce
  a syntactic constraint. And a syntactic constraints may be not
  joined to a specific concept.

\item \verb#print <expr>;# Displays the expression at standard output
  (useful for debugging).

\item \verb#println <expr>;# Does the same and displays a new line.
  
\item  \verb#<expr_1> = <expr_2>;#
  
  An assignment of the expression $<expr_1>$ with the value of the
  expression $<expr_2>$. The expression $<expr_1>$ is a complex
  variable, or simple variable, the expression $<expr_2>$ is a constant
  or a complex variable or a simple variable such that there is a
  environment \footnote{An environment is an application $\phi$ from
    $E$ to $F$ such that $E$ is a set of variables and $F$ a set of
    constants or variables. The resolution of a variable $\alpha$ in
    an environment is $\beta = \phi^k(\alpha)$ such that $\beta$ is a
    constant. It should be noted that \texttt{Elvex} also handles
    complex variables. In this case $\phi^k(\alpha)$ is replacing
    variables in terms by their constants. } of the syntagmatic rule
  to solve it.
    
The assignments of $\uparrow$ and $\Downarrow_i$ are not
  possible, as these attributes are implicitly assessed during the
  text generation and modification would result in
  inconsistencies.

    The assignments are limited to these cases:
    \begin{itemize}
      
    \item Assign a list
      
      \begin{itemize}
      \item  \verb#<#$\dots$\verb#> = <#$\dots$\verb#># Assigns a list to
        another one.
      \item \verb#<#$\dots$\verb#> = $X# Assigned by the value of the
        variable \verb#$X#
      \end{itemize}
      
    \item Assign an inherited attribute
    
      \begin{itemize}
      \item $\downarrow$\verb#i = $X# Assigned by the value of the
        variable \verb#$X#
      \item $\downarrow$\verb#i = [#$\dots$\verb#]# Assigned by a
        constant or variable feature structure.
      \item $\downarrow$\verb#i = #$\uparrow$ Assigned by an inherited attribute.
      \item $\downarrow$\verb#i = #$\dots$\verb# #$\cup$\verb# #$\dots$
        Assigned by the result of the unification of two expressions
        (feature structures or attributes).
      \item $\downarrow$\verb#i = #$\Downarrow$\verb#j# Assigned by a
        synthesized attribute.
      \end{itemize}
      
    \item Assign a synthesized attribute
    
      \begin{itemize}
      \item $\Uparrow$\verb# = $X#
      \item $\Uparrow$\verb# = [#$\dots$\verb#]#
      \item $\Uparrow$\verb# = #$\uparrow$
      \item $\Uparrow$\verb# = #$\dots$\verb# #$\cup$\verb# #$\dots$
      \item $\Uparrow$\verb# = #$\Downarrow$\verb#j#
      \end{itemize}
      
    \item  Assign a simple variable
    
      \begin{itemize}
      \item \verb#$X = $Y#
      \item \verb#$X = a# Assigned by a constant or a literal.
      \item \verb#$X = <#$\dots$\verb#>#
      \item \verb#$X = [#$\dots$\verb#]#
      \item \verb#$X = #$\uparrow$
      \item \verb#$X = #$\dots$\verb# #$\cup$\verb# #$\dots$
      \item \verb#$X = #$\Downarrow$\verb#j#
      \item \verb#$X = <expr>#  by evaluating an arithmetical or logical expression
      \item \verb#$X = <search># by evaluating a lexical entry (see below)
      \end{itemize}
      
    \item Subsume a feature structure
    
      A feature structure is subsumed

      \begin{itemize}
      \item \verb#[#$\dots$\verb#] #$\subset$\verb# #$\uparrow$ by an inherited attribute.
      \item \verb#[#$\dots$\verb#] #$\subset$\verb# #$\Downarrow$\verb#j# by a synthesized attribute
      \item \verb#[#$\dots$\verb#] #$\subset$\verb# $X# by the value of a variable.
      \item \verb#[#$\dots$\verb#] #$\subset$\verb# <search># by a lexical entry
      \end{itemize}
      
    %\item \verb#foreach <variable> in <list> <rules>#

      %Interprets operational semantics \verb#<rules># foreach element of \verb#list#.

    \item \verb#if (<boolExpr>) <rules>#
    
      Interprets operational semantics \verb#<rules># if and only
      if the expression \verb#<boolExpr># is verified.

    \item \verb#if (<boolExpr>) <rules_1> else <rules_2>#
      
      Interprets operational semantics \verb#<rules_1># if and
      only if \verb#<boolExpr># is verified, otherwise interpret \verb#<rules_2>#.
      
    \end{itemize}

    A logical expression \verb#<booExpr># is

    \begin{itemize}
    \item \verb#<boolExpr>#  $\lor$ \verb#<boolExpr># Or
    \item \verb#<boolExpr>#  $\land$ \verb#<boolExpr># And
    \item $\neg$ \verb#<boolExpr># Negation
    \item \verb#<boolExpr>#  $\Rightarrow$ \verb#<boolExpr>#
      Material implication
    \item \verb#<boolExpr>#  $\Leftrightarrow$ \verb#<boolExpr>#
      Biconditional
    \item \verb#<expr>#  $==$ \verb#<expr># Equality
    \item \verb#<expr>#  $\neq$ \verb#<expr># Difference
    \item \verb#<expr># Expression evaluated as Boolean: A
      feature structure is evaluated as true if it is not
      \verb#NIL# or $\bot$ (result of the failure of unification
      between two feature structures).
    \end{itemize}
    
    The expressions are:
    
    \begin{itemize}
    \item Integer evaluated as true if not zero
    \item Double evaluated as true if not zero
    \item String evalued as true if not empty
    \item \verb#<expr>#  $\cup$ \verb#<expr># Unification of two
      feature structures
    \item $\uparrow$
    \item $\Uparrow$
    \item $\downarrow_i$
    \item $\Downarrow_i$
    \item \verb#[...]# Feature structure
    \item \verb#NIL# Null value of a feature
    \item \verb#a|b...# Constant evaluated as true
    \item \verb#$X# Variable
    \item \verb#<...># List evaluated as true if not null (\verb#<>#)
    \end{itemize}
    
Let's add algebraic expressions and numbers:
    
    \begin{itemize}
    \item \verb#<expr>#  $<$ \verb#<expr>#
    \item \verb#<expr>#  $\le$ \verb#<expr>#
    \item \verb#<expr>#  $>$ \verb#<expr>#
    \item \verb#<expr>#  $\ge$ \verb#<expr>#
    \item \verb#<expr>#  $+$ \verb#<expr>#
    \item \verb#<expr>#  $-$ \verb#<expr>#
    \item \verb#<expr>#  $*$ \verb#<expr>#
    \item \verb#<expr>#  \% \verb#<expr>#
    \item \verb#<expr>#  $/$ \verb#<expr>#
    \item $-$ \verb#<expr>#
    \item Double-precision floating-point format number
    \end{itemize}
    
  \end{itemize}
  
\subsubsection*{Variable evaluation}

Variables are evaluated locally at a syntagmatic rule. This
defines the scope of a variable.

Here are different methods for solving variables:

\begin{itemize}
\item Direct assignment
\begin{lstlisting}
S → S {
  [vform:NIL];
  [tense:$tense, aspect:$aspect, mode:$mode] ⊂ ↑;
  if (!$tense)
    $tense = present;
...
}
\end{lstlisting}
Line 5 indicates that the value of the variable \verb#$tense# is
\verb#present# by default (in the absence of another value previously given). The
value of
\verb#$tense# is eventually given by the inherited attribute
$\uparrow$ online 3.

Note that the order of evaluation of the rules is indifferent and that it
is not necessary to precede rule (3) in the order that they are listed.

\item Guard
\begin{lstlisting}[numbers=none]
S → NP VP {
  [subject:$inheritedSubject, $Rest];
...
}
\end{lstlisting}

The value of the variable \verb#$inheritedSubject# is given 
by the $uparrow$ subject feature. The value of the variable 
\verb#$Rest# is the feature structure that remains when you remove the
feature \verb#subject#.

\item Subsumption
\begin{lstlisting}[numbers=none]
S → NP VP {
...
  [subject:$synthesizedSubject] ⊂ ⇓2;
...
}
\end{lstlisting}

The value of the variable \verb#$synthesizedSubject# is given 
by the \verb#subject# feature of $\Downarrow2$. 

\item Binding a variable within a syntagmatic rule
\begin{lstlisting}[numbers=none]
S → NP VP {
  [subject:$inheritedSubject, $Rest];
  [subject:$synthesizedSubject] ⊂ ⇓2;
  ↓1 = [subject:$inheritedSubject] 
  		∪ [subject:$synthesizedSubject];
...
}
\end{lstlisting}

The variables \verb#$synthesizedSubject# and \verb#$inheritedSubject#
are both used within this rule to be
transmitted to the \verb#subject# feature of the inherited attribute $\downarrow1$.

\item Linking a variable within a lexical entry

\begin{lstlisting}[numbers=none]
a	v[HEAD:brouillard, subjC:[FORM:"il"], 
		objC:[HEAD:brouillard, det:yes, 
		def:no, part:yes, neg:$Neg], 
		locCl:[HEAD:_pro, number:sg, 
			person:three], 
		vform:tensed, vtense:present, 
		mode:indicative, 
		subj:[person:three, number:sg], 
		neg:$Neg, fct:none]
\end{lstlisting}

The variable \verb#$Neg# is linked in this lexical entry to the expression "it is foggy" so that its use as negative propagates a negation of the noun:

\begin{exe}
\ex \begin{xlist} 
\ex \label{0-a} \textit{Il y a du brouillard} (it is foggy)
\ex \label{0-a} \textit{Il n'y a pas de brouillard} (There is no fog)
\ex \label{0-a} \textit{*Il y a de brouillard} ()
\ex \label{0-a} \textit{*Il n'y a pas du brouillard}
\end{xlist}
\end{exe}

\end{itemize}

\subsection{Lexicon}

The lexicon is a list of entries separated by semicolons.

Each entry defines the sequence to be produced, followed by a disjunction
(noted by the character \verb#|#) of lexical categories
optionally followed by a feature structure.

The sequence to be produced may be:

\begin{itemize}
\item A string of characters between \verb#""#, in which case the sequence
  will be exactly this one. Note that the sequence may be empty.

\item A string of characters between \verb#""# containing
  variables. The sequence produced will be this one with the variables
  substituted by their respective values.

\item \texttt{FORM}: the sequence produced will literally be the one
  given by the feature \texttt{FORM}. \texttt{Elvex} version 2.10 only provides for
  constant, subsequent versions will also have
  complex variables.

\end{itemize}

The category is an identifier, namely the final symbol of grammar.

The feature structure allows to represent all the
constraints imposed by the lexicon.

A few examples:

\begin{itemize}
\item \verb#"?" interrogativeDot;#

\bigskip
\item \verb#"tag-$Line" tag[line:$Line]; #

\bigskip
\item \verb# FORM title; #

\bigskip
\item \verb#"elles-mêmes"	itself [gender:fm, number:pl, person:three, itself:yes]; #

\bigskip
\item \begin{verbatim}
serai	aux_être [aux:être, voice:active, finite:yes,
  mode:indicative,  vtense:future_anterieur, vform:tensed, 
subj:[person:one, number:sg]]; 
\end{verbatim}

\bigskip
\item \begin{verbatim}
// Lexical input "temperature" producing 
//"La température est de xxx degrés" (the temperature is xxx degrees Celsius"
est v[HEAD:température, 
	subjC:[HEAD:température, number:sg, det:yes, def:yes], 
	pobjC:[HEAD:degré, number:pl, pcas:de, det:yes,
		detNum:[HEAD:num, value:$Deg]], 
	vform:tensed, vtense:present, mode:indicative, 
	subj:[person:three, number:sg], fct:none, value:$Deg]

//"la température est douce" (the temperature is mild)
|v[HEAD:température, subjC:[HEAD:température, number:sg, 
	det:yes, def:yes], modC:[HEAD:doux, number:sg, gender:fm],
	vform:tensed, vtense:present, mode:indicative, 
	subj:[person:three, number:sg], fct:few]

//"la température est élevée" (the temperature is high)
|v[HEAD:température, subjC:[HEAD:température, number:sg, 
	det:yes, def:yes], modC:[HEAD:élevé, number:sg, gender:fm],
	vform:tensed, vtense:present, mode:indicative, 
	subj:[person:three, number:sg], fct:much];
\end{verbatim}

\bigskip
\item \begin{verbatim}
// Lexical entry of "temperature".
// producing "Il fait chaud" (It's hot)
fait	v[HEAD:température, subjC:[FORM:"il"], 
	modC:[HEAD:chaud, number:sg, gender:ms], vform:tensed,
	vtense:present, mode:indicative, 
	subj:[person:three, number:sg], neg:$Neg, fct:high]

//"Il fait très chaud" (It's very hot)
|v[HEAD:température, subjC:[FORM:"il"], modC:[HEAD:chaud, 
	number:sg, gender:ms, mod:<[HEAD:très]>], vform:tensed, 
	vtense:present, mode:indicative, subj:[person:three, 
	number:sg], neg:$Neg, fct:very_high]

//"Il fait très froid" (It's very cold)
|v[HEAD:température, subjC:[FORM:"il"], modC:[HEAD:froid, 
	number:sg, gender:ms, mod:<[HEAD:très]>], vform:tensed,
	vtense:present, mode:indicative, subj:[person:three, 
	number:sg], neg:$Neg, fct:very_low]

//"Il fait froid" (It's cold)
|v[HEAD:température, subjC:[FORM:"il"], modC:[HEAD:froid, 
	number:sg, gender:ms], vform:tensed, vtense:present, 
	mode:indicative, subj:[person:three, number:sg], 
	neg:$Neg, fct:low]; 
\end{verbatim}
	
\bigskip
\item \begin{verbatim}
// Lexical entry of "rain" producing
// "Il y a une pluie fine" (A fine rain fell)
a	v[HEAD:pleuvoir, subjC:[FORM:"il"], 
	objC:[HEAD:pluie, det:yes, number:sg, qual:yes,
		def:no, mod:<[HEAD:fin, pos:post]>], 
	locCl:[HEAD:_pro, number:sg, person:three], 
	vform:tensed, vtense:present, mode:indicative, 
	subj:[person:three, number:sg], neg:$Neg, fct:few]
	
// "il y a un peu de pluie" (It is raining a bit)
|v[HEAD:pleuvoir, subjC:[FORM:"il"], 
	objC:[HEAD:pluie, det:yes, number:sg, 
	detForm:[FORM:"un peu de"]], 
	locCl:[HEAD:_pro, number:sg, person:three], 
	vform:tensed, vtense:present, mode:indicative, 
	subj:[person:three, number:sg], neg:$Neg, fct:not_much]

// "il y a beaucoup de pluie" (There is plenty of rainfall)
|v[HEAD:pleuvoir, subjC:[FORM:"il"], 
	objC:[HEAD:pluie, det:yes, number:sg, 
	detForm:[FORM:"beaucoup de"]], 
	locCl:[HEAD:_pro, number:sg, person:three], 
	vform:tensed, vtense:present, mode:indicative, 
	subj:[person:three, number:sg], neg:$Neg, fct:much]

// "il y a de fortes précipitations" (It is raining so much)
|v[HEAD:pleuvoir, subjC:[FORM:"il"], 
	objC:[HEAD:précipitation,
        det:yes, number:pl, qual:yes, def:no, mod:<[HEAD:fort,
        pos:pre]>], locCl:[HEAD:_pro, number:sg, person:three],
        vform:tensed, vtense:present, mode:indicative,
        subj:[person:three, number:sg], neg:$Neg, fct:a_lot]; 
\end{verbatim}

\end{itemize}

\subsection{Input}

The input is summarized by the type of phrase or text you want to generate, followed by a
feature structure containing the concept and the illocutionary structure to be expressed.


\texttt{HEAD} allows you to define the named concept to be generated.

examples

\begin{itemize}
\item 
\begin{verbatim}
sentence [HEAD:raining, neg:yes]
\end{verbatim}

Simple entry for the concept of \textit{raining} producing 
\begin{exe}
\ex \begin{xlist} 
\ex \textit{Il ne pleut pas} (It is not raining)
\ex \textit{Il n'y a pas de pluie} (There is no rain)
\end{xlist}
\end{exe}


\item 
\begin{verbatim}
sentence 
   [HEAD:to_ask,
    i:[HEAD:menuisier, id:3, number:sg, gender:ms, def:yes],
    iii:[HEAD:apprenti, id:6, def:yes, gender:ms, number:sg],
    ii:[HEAD:scier,
        i:[idref:6],
        ii:[HEAD:poutre, number:sg, def:yes]
        ]
    ]
\end{verbatim}

Complex concept where the features \texttt{id} and \texttt{idref}
make it possible to constrain co-references. 

This entry produces 
\begin{exe}
\ex 
\begin{tabbing}
\textit{Le menuisier} \= \textit{demande} \= \textit{à} \textit{l'apprenti}~~~ \= \textit{de} \textit{scier} \= \textit{la} \textit{poutre}\\
The carpenter \> asks \> the apprentice \> to saw \> the beam\\
The carpenter asks the apprentice to saw the
beam
\end{tabbing}
\end{exe}

\item 
\begin{verbatim}
sentence 
   [HEAD:raining, 
    tense:future, 
    modSType:time, 
    modS:<[FORM:"lundi 18 mars 2019", type:time]>
   ]
\end{verbatim}

Concept producing an adverbial on the basis of the
constant "lundi 18 mars 2019" (Monday, the 18th of March 2019). This entry produces:

\begin{exe}
\ex \textit{Il va pleuvoir le lundi 18 mars 2019.} (It will rain on Monday, March 18, 2019) 
\end{exe}

\item 
\begin{verbatim}
sentence 
   [HEAD:cause, 
    i: [HEAD:raining]]
    ii: [HEAD:se_couvrir, i:[HEAD:you], modality:devoir]
   ]
\end{verbatim}

Input for the following generated sentences:
\begin{exe}
\ex \begin{xlist} 
\ex \textit{Il pleut, tu dois te couvrir} (It is raining, you have to dress warmly)
\ex \textit{Il pleut, par conséquent tu dois te couvrir} (It is raining, therefore you have to dress warmly)
\ex \textit{Tu dois te couvrir s'il pleut} (You have to dress warmly if it is raining)
\ex \textit{Tu dois te couvrir car il pleut} (You have to dress warmly because it is raining)
\ex \textit{S'il pleuvait, tu devrais te couvrir} (If it rained, you should dress warmly)
\ex \textit{S'il pleut, tu devras te couvrir} (If it is raining, you should dress warmly)
\ex \textit{Il pleuvrait, tu devrais te couvrir} (It would rain, you should dress warmly)
\end{xlist}
\end{exe}

\item 
\begin{verbatim}
sentence 
   [HEAD:initiative, 
    i:[HEAD:Jean], 
    objC:[lexFct:Magn, mod:<[HEAD:audacieux]>], 
    lexFct:Oper1
   ]
\end{verbatim}

Input containing features to fix lexical functions. The
production is 
\begin{exe}
\ex \textit{Jean prend une initiative très audacieuse.} (Jean takes a very bold initiative)
\end{exe}

\item 
\begin{verbatim}
sentence 
   [HEAD:initiative, 
    i:[HEAD:Jean], 
    objC:[lexFct:Magn], 
    lexFct:Oper1
   ]
\end{verbatim}

Input for

\begin{exe}
\ex \textit{Jean prend une belle initiative.}  (Jean takes a fine initiative)
\end{exe}

\item 
\begin{verbatim}
sentence 
   [HEAD:to_break,
    i:[HEAD:Jean], 
    ii:[HEAD:carafe, number:sg, def:yes, 
       mod:<[HEAD:little],[HEAD:white]>], 
    diathesis:passive, tense:past, neg:yes, 
    modV:<[HEAD:roughly]>
   ]
\end{verbatim}

Input containing a negative passive feature. The
production is
\begin{exe}
\ex \textit{La petite carafe blanche n'avait pas été cassée
brutalement par Jean.} (The small white carafe had not been broken
brutally by Jean.)
\end{exe}

\end{itemize}

\section{Design pattern (to be completed)}

A design pattern applies in a general case which can be summarized by
some constraints. It is not associated with a specific event, but to a
recurring organization of the grammar for which a single and coherent
answer must be provided.

\subsubsection*{A - Local propagation of synthesized attributes}

Context:
A feature \texttt{A} that must constrain a result \texttt{R}, where
\texttt{R} depends only on the lexicon or grammar.

Examples of applications:
\begin{itemize}
\item Production of a constrained noun in an idiomatic phrase
  (\textit{breaking the \textbf{ice}})
\item Conjugation to the passive voice
\end{itemize}

The general rule is written as follows

\begin{lstlisting}
X → Y {
  [A:$v, $rest];
  [ARealization:R] ⊂ ⇓1;
  ↓1 = [R:$v, $rest];
}
\end{lstlisting}

Example 1: The sentence \textit{We basically got taken to the cleaners
  on that deal.} introduces the fixed noun \textit{cleaners} which
comes from the verb \textit{take one to the cleaner}. Here the agent of the
passive sentence is not represented.

\begin{lstlisting}
VN → V [PP] PP {
  [obj:NIL, obl:$Obl];
  [subj:$subj] ⊂ ↑;
  ↓1 = ↑;
  ↓3 = ↑$Obl;
  [oblC:$oblSynt] ⊂ ⇓1;
  if (#2) {
    attest $oblSynt;
    ↓2 = $oblSynt;
  }
  else {
    attest ¬$oblSynt;
  }
}

taken verb[HEAD:take_one_to_the_cleaners, 
	constI:[HEAD:cleaner, number:pl, def:yes]];
\end{lstlisting}

Example 2:
The agent of a clause is carried out by the grammatical function subject
for conjugation to the active voice, and by the oblique function in
\textit{by} for a passive voice conjugation.

\begin{lstlisting}
// agent as subject
Sentence → Sentence {
  [agent:$agent, $rest];
  [agentRealization:subject] ⊂ ⇓1;
  ↓1 = [subject:$agent, $rest];
  ⇑ = ⇓1;
}

// agent as oblique
Sentence → Sentence {
  [agent:$agent, $rest];
  [agentRealization:oblBy] ⊂ ⇓1;
  ↓1 = [byObj:$agent, $rest];
  ⇑ = ⇓1;
}

// patient as object
Sentence → Sentence {
  [patient:$patient, agent:NIL, $rest];
  [patientRealization:object] ⊂ ⇓1;
  ↓1 = [object:$patient, $rest];
  ⇑ = ⇓1;
}

// patient as subject
Sentence → Sentence {
  [patient:$patient, agent:NIL, $rest];
  [patientRealization:subject] ⊂ ⇓1;
  ↓1 = [subject:$patient, $rest];
  ⇑ = ⇓1;
}

// passive voice
Sentence → NP VERB PP {
  [subject:$subject, parObl:$parObl, $rest];
  ↓1 = $subject;
  ↓2 = $rest ∪ [voice:passive];
  ↓3 = $parObl;
  ⇑ = [agentRealization:oblBy, 
	patientRealization:subject];
}

// active voice
Sentence → NP VERB NP {
  [subject:$subject, object:$object, $rest];
  ↓1 = $subject;
  ↓2 = $rest ∪ [voice:active];
  ↓3 = $object;
  ⇑ = [agentRealization:subject, 
	patientRealization:object];
}
\end{lstlisting}

\subsubsection*{B - Transformation of the attribute inherited by a syntactic head}

Context:
An inherited attribute \texttt{A} that must constrain a realization
\texttt{R} by a new predicate in generation.

Examples of applications:
\begin{itemize}
\item Use of a verbal tense auxiliary or a modal auxiliary
\item Compound conjugation in French
\end{itemize}

The general rule is written as follows

\begin{lstlisting}
X → Y {
  [A:K];
  [HEAD:$HEAD, A:k, $Rest] ⊂ ↑;
  ↓1 = [HEAD:L, arg:[HEAD:$HEAD, $Rest]];
}

\end{lstlisting}

Example:
In French, the aspect is reflected in conjugated forms, but also in
compound verb constructions (\og être en train de + inf\fg{} (to be in the
process of doing sth). 

\begin{lstlisting}
sentence → sentence {
  [aspect:progressive];
  [HEAD:$HEAD, aspect:progressive, i:$I, $Rest] ⊂ ↑;
  ↓1 = [HEAD:être_en_train_de, 
	i:[HEAD:$HEAD, i:$I, $Rest]];
  ⇑ = ⇓1;
}
\end{lstlisting}

\section{The \texttt{elvex} command}

\texttt{Elvex} is an Natural Text Generator. It takes as an input a
concept, a local lexicon, a compacted lexicon, and a grammar, and then
outputs a text corresponding to the concept writing in Natural
Language.

\subsection*{Compilation and full installation of \texttt{Elvex}}

The full compilation of \texttt{Elvex} requires the following programs on your machine:

\begin{itemize}
\item g++
\item flex
\item bison
\item cmake
\item libxml2-dev
\end{itemize}

The full command is:

\begin{lstlisting}[frame=none, numbers=none]
cmake .
make
sudo make install
\end{lstlisting}

\subsubsection{try \texttt{Elvex}}

\begin{lstlisting}[frame=none, numbers=none]
. ./try-me.sh
\end{lstlisting}

\subsection{\texttt{Elvex} command}

The \texttt{elvex} command is:\\

\noindent
\verb#elvex [options] [input]*#\\

The options are:

\begin{description}
\item \verb#-h|--help#

Displays help and exits

\item \verb#-v|--version#

Displays the version number and exits

\item \verb#-a|--reduceAll#

  Reduces all rules during the generation process. This mode
  allows fine debugging and requires an understanding of the
  generation process to be mastered.

\item \verb#-t|--trace#

  Displays the reduced rules during the generation process. This mode allows one for a
  grammar debugging.

\item \verb#-r|--random#
  
  Displays only one randomly selected output.
  
\item \verb#-o|--one#
  
  Displays only the first output.
  
\item \verb#-rulesFile <file>#
  
  File containing the rules.
  
\item \verb#-lexiconFile <file>#
  
  File containing the lexicon. 

\item \verb#-inputFile <file>#
  
  File containing the input. 
  
\item \verb#-compactDirectory <directory>#

  Directory containing the files of the compacted lexicon.

\item \verb#-compactLexiconFile <filePrefix>#

  Prefix of the compacted lexicon files
  \verb#<directory>/<filePrefix>.tbl# and
  \verb#<directory>/<filePrefix>.fsa#
  
\item \verb#-maxLength <integer>#

  Defines the maximum length of the sequence to be produced.

\item \verb#-maxUsages <integer>#

  Defines the maximum number of times the same rule can be used.

\item \verb#-maxItems <integer>#

  Defines the maximum cardinality of the item sets.

\item \verb#-maxTime <integer>#

  Defines the number of seconds before the program stops.

\end{description}

\subsection{Grammar debugging}

\noindent
\verb#elvexdebug [options] [input]*#\\

The \texttt{elvexdebug} command works the same way with same options
but produces a \texttt{HTML} document on the standard output. You must
use your favorite browser to view this document properly and review
the computation.

The command \texttt{elvexdebug} is recommended to set the grammars in
the most difficult cases.

Where the specific options are:

\begin{description}

\item \verb#--traceInit#

  Trace the initial items
  
\item \verb#--traceStage#

  Trace the items sets for each generation stage
  
\item \verb#--traceClose#

  Trace the items sets closure
  
\item \verb#--traceShift#

  Trace the items for each lexical entry
  
\item \verb#--traceReduce#

  Trace the items for each reduction

\item \verb#--traceAction#

  Trace the operational semantics results
  
\end{description}

\subsection{\texttt{elvexbuildlexicon} for managing big lexicons}

The command of \texttt{ElvexBuildLexicon} is

\noindent
\verb#elvexbuildlexicon [global-option] <build|consult|list> <input>#\\

Where the options are:

\begin{description}
\item \verb#-h|--help#

Displays help and exits

\item \verb#-d <directory>#

  Directory containing the files of the compacted lexicon.

\item \verb#-f <filePrefix>#

  Prefix of the compacted lexicon files
  \verb#<directory>/<filePrefix>.tbl# and
  \verb#<directory>/<filePrefix>.fsa#.
  
\item \verb#-patternFile <file>#

  The pattern file which consists of lines
  \verb#lexeme	pos	lemma	features#

\item \verb#-morphoFile <file>#
  
  The morphological file which consists of lines
  \verb#form	pos	lemma	features#

\end{description}

The lexicon is the mapping of forms where the features of pattern and
morpho are unified for each couple of (pos, lemma).

\subsubsection*{Command to build a compacted lexicon}

\begin{description}
\item \verb#$elvexbuildlexicon -d <directory> -f <filePrefix> build <input># 
\end{description}

\subsubsection*{Command to test a compacted lexicon}

\begin{description}
\item \verb#$elvexbuildlexicon -d <directory> -f <filePrefix> list# 

\subsubsection*{Command to consult a compacted lexicon}

\item \verb#$elvexbuildlexicon -d <directory> -f <filePrefix> consult# 

\end{description}

The compacted lexicon is a set of lexical entries of the type

\paragraph{}\verb@pos#lexeme	form#FS|form#FS...@

\begin{itemize}
\item \verb#pos# is a part of speach
\item \verb#lexeme# is the lexeme given by the \verb#HEAD# feature
\item \verb#form# is the produced form (a string of UTF8 characters in version 2.10)
\item \verb#FS# is a feature structure
\end{itemize}

\end{document}
